PIC simulation of instabilities in relativistic supernovae shocks

### 2021-04-16 papers/Physica2021/Romansky_Physica2021.tex
paper

	\begin{abstract}
                 Modern observations show presence of relativistic supernovae - objects intermediate between regular supernovae and gamma-ray bursts. The typical lorentz-factors of plasma flows in relativistic supernovae are about $1.5$. The process of cosmic ray acceleration is very different from both non-relativistic and ultra-relativistic case. In this paper is presented particle-in-cell modeling of trans-relativistic collisionless shocks. It is shown, that development of magneto-hydrodynamic instabilities ii electron-ion relativistic shock, particularly Bell's instability, has strong influence on cosmic ray acceleration. At times larger than $1500$ reverse ion plasma frequency magnetic field, generated by Bell's instability, significantly suppresses the electrons spectrum. This effect should be taken into account for modeling of synchrothron spectrum from relativistic supernovae.
	\end{abstract}

	\section{Introduction}
	Relativistic shocks are widely discussed phenomena in modern astrophysicist. They can be generated in the supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting supermassive black holes \cite{1984RvMP...56..255B}, the stellar masses black holes \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}. Relativistic shocks in astrophysical objects are sources of cosmic rays \cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P} and can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.

	The presence of high energy non-thermal particles leads to development of various magneto-hydrodynamic instabilities. Bell instability \cite{Bell04} can be the most significant due to it's large growth rate. Instabilities modify the flow and has strong influence on process of particle acceleration. Below we present results of particle-in-cell modeling of relativistic shock and show that growth of magnetic field due to instabilities decreases efficiency of electron acceleration on shock front.

@@

	Inclination angle of magnetic field due to shock velocity is $\theta = 30$ degrees, which corresponds to the maximum efficiency of particle acceleration \cite{Romansky18} and magnetic field is lying in the simulation plane.

@@

	Profiles of transversal magnetic field, obtained from numerical simulation, show significant field amplification in precursor of shock wave, see figure \ref{field}. It is result of development of instabilities in plasma, particularly Bell instability \cite{Bell04}, which growth rate is larger then growth rate of other instabilities.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.8\textwidth]{fig/Bnorm.png} 
		\caption{Evolution of the magnetic field in time.}
		\label{field}
	\end{figure}

	Spectrum of electrons is shown in figure \ref{spectrume}. One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1500 inverse protons plasma frequencies and at the longer times spectrum decreases. It may by explained as consequence of magnetic field amplification. Instabilities increase the transverse magnetic field and also the angle between field and shock velocity. As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011}. On the other hand, protons distribution function constantly increases in time, as shown in figure \ref{spectrump}. We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses he electrons injection in acceleration process.

	\begin{figure}[h!]
		\centering
		\begin{minipage}{0.8\textwidth}
			\centering
			\includegraphics[width=0.98\textwidth]{fig/spectrum.png} 
			\caption{Temporal evolution of electrons spectrum.}
			\label{spectrume}
		\end{minipage}\hfill
		\begin{minipage}{0.8\textwidth}
			\centering
			\includegraphics[width=0.98\textwidth]{fig/spectrump.png} 
			\caption{Temporal evolution of protons spectrum.}
			\label{spectrump}
		\end{minipage}
	\end{figure}

@@

	Particle-in-cell simulations have been used to derive effective macroscopic parameters of relativistic shocks in space plasmas. The effective adiabatic index and electron temperatures are two important parameters required to supplement large-scale single-fluid magneto-hydrodynamic simulations of processes in astrophysical plasmas. The adiabatic indexes for the downstream flow derived with PiC method are summarized in Table 1 and can be used for MHD modeling of relativistic
	supernovae. Further modelling is needed to study the dependence of these two parameters on the initial conditions which, reflect the presence of MHD large-scale turbulence in the stellar wind of the progenitor massive star upstream of the  mildly-relativistic shock of the supernova.

>>>>

	\begin{abstract}
                 Radio observations revealed a presence of relativistic supernovae a class of objects intermediate between regular supernovae and gamma-ray bursts. The typical Lorentz-factors of plasma flows in relativistic radio-bright supernovae were estimated to be about 1.5. Mildly relativistic shocks in electron-ion plasmas are known to efficiently accelerate radio-emitting electrons if the shock is subluminous. The inclination angle of the subluminous shock to the ambient magnetic field should be below a critical angle which depends on the Mach number and the plasma magnetization parameter.  In this paper we present particle-in-cell modeling of electron acceleration by a subluminal mildly-relativistic collisionless shock. It is shown that  a development of the ion scale Bell-type instability in electron-ion relativistic shock may have a strong influence on the electron injection and acceleration. In time period of about $1500 \omega_{pi}^{-1}$  ($\omega_{pi}$ is the ion plasma frequency) after the shock initialization the  magnetic field fluctuations  generated by Bell's instability may significantly suppresses the electrons spectrum even in a sub-luminous shock. This effect is important to account for modeling of synchrothron spectra from relativistic supernovae.
	\end{abstract}

	\section{Introduction}
	Relativistic shocks are widely discussed phenomena in modern astrophysics. They can be generated in the supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting supermassive black holes \cite{1984RvMP...56..255B}, the stellar masses black holes \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and pulsar wind nebulae \cite{2019MNRAS.488.5690O,2017SSRv..207..235B,2017JPlPh..83e6301K,2019ApJ...876L...8B}. Relativistic shocks in astrophysical objects are sources of cosmic rays \cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P} and can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}.

	The presence of high energy non-thermal particles leads to development of various magneto-hydrodynamic instabilities. Bell instability \cite{Bell04} can be the most significant due to it's large growth rate. Instabilities modify the flow and has strong influence on process of particle acceleration. We present below results of particle-in-cell modeling of relativistic shock and show that growth of magnetic field due to instabilities decreases efficiency of electron acceleration on shock front.

@@

	Inclination angle of magnetic field due to shock velocity varies in different setups from $\theta = 20$ degrees to $\theta = 40$. This values are chosen to contain critical angle, which is defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, and is equal approximately $30$ degrees. If inclination angle is larger then critical value, shock becomes superluminal - particle can not escape from it to upstream, even moving along the field lines with speed of light \cite{Pelletier2017,Sironi2011}. The maximum efficiency of particle acceleration is near the critical angle, less below it \cite{Romansky18}. Magnetic field is lying in the simulation plane.

@@

	Profiles of transversal magnetic field, obtained from numerical simulation, show significant field amplification in precursor of shock wave, see figure \ref{field}. It is result of development of instabilities in plasma, particularly Bell's instability \cite{Bell04}, which growth rate is larger then growth rate of other instabilities.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.8\textwidth]{fig/Bnorm.png} 
		\caption{Temporal evolution of the transversal magnetic field.}
		\label{field}
	\end{figure}

	It is known, that development of Bell's instability can suppress electrons acceleration \cite{Crumley2019}. We study this phenomena with different inclination angles of magnetic field. Spectrum of electrons is shown in figure \ref{spectrume}. One can see, that electrons distribution function at high energies $(\gamma > 300)$ is maximal at times about 1400 inverse protons plasma frequencies and at the longer times spectrum decreases. It may by explained as consequence of magnetic field amplification. Instabilities increase the transverse magnetic field and also the angle between field and shock velocity. As a result shock becomes quasi-perpendicular and can not effectively accelerate particles \cite{Sironi2011,Romansky18}. On the other hand, protons distribution function constantly increases in time, as shown in figure \ref{spectrump}. We assume, that oscillating transverse field has the strongest influence on the smallest scales, and suppresses the electrons injection into acceleration process.

	\begin{figure}[h!]
		\centering
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.95\textwidth]{fig/spectrum20.png} 
		\end{minipage}\hfill
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.95\textwidth]{fig/spectrum30.png} 
		\end{minipage}
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.95\textwidth]{fig/spectrum33.png} 
		\end{minipage}
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.95\textwidth]{fig/spectrum40.png} 
		\end{minipage}
		\caption{Temporal evolution of electrons spectrum in shocks with different inclination angles. Top row: left $\theta = 20^\circ$, right $\theta = 25^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
		\label{spectrume}
	\end{figure}

\begin{figure}[h!]
	\centering
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{fig/spectrump20.png} 
	\end{minipage}\hfill
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{fig/spectrump30.png} 
	\end{minipage}
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{fig/spectrump33.png} 
	\end{minipage}
	\begin{minipage}{0.49\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{fig/spectrump40.png} 
	\end{minipage}
	\caption{Temporal evolution of protons spectrum in shocks with different inclination angles. Top row: left $\theta = 20^\circ$, right $\theta = 25^\circ$, bottom row: left $\theta = 33^\circ$, right $\theta = 40^\circ$}
	\label{spectrump}
\end{figure}

@@

	Details of electrons acceleration are important for correct interpretations of observations synchrothron radiation from astrophysical sources. Particle-in-cell simulation has been used to demonstrate magnetic instabilities influence on electrons acceleration in relativistic electron-ion shock. It is shown, that amplification of magnetic field suppresses formation of electrons spectrum at high energies.This effect should be taken into account while deriving parameters of source of radiation via analysis of synchrothron spectrum. Further PIC simulations with a wider spatial and energetical ranges will allow to study the effect more precisely.
